## Example: HERS dataset {#sec-hers-lm-example}

:::{.callout-note}
This example is based on @vittinghoff2e, Chapter 4.
:::

::: notes

{{< include _sec_hers_intro.qmd >}}

:::

### Load the data

```{r}
#| eval: false
#| code-fold: show
library(haven)
hers <- haven::read_dta(
  paste0(
    "https://regression.ucsf.edu/sites/g/files",
    "/tkssra6706/f/wysiwyg/home/data/hersdata.dta"
  )
)
```

```{r}
#| include: false
hers <- fs::path_package("rme", "extdata/hersdata.dta") |>
  haven::read_dta()
```

:::{#tbl-hers-data-lm}

```{r}
hers |> head()
```

First six rows of the `hers` dataset

:::

### Research question

::: notes
What is the cross-sectional association between
body mass index (BMI) and LDL cholesterol
among women with coronary heart disease,
adjusting for potential confounders?
:::

### Simple linear regression

::: notes
We begin with a simple linear regression of LDL cholesterol on BMI,
following @vittinghoff2e Table 4.1.
:::

$$E[\text{LDL} \mid \text{BMI} = b] = \beta_0 + \beta_1 b$$ {#eq-hers-simple}

#### Exploratory plot

:::{#fig-hers-ldl-bmi}

```{r}
library(ggplot2)
hers |>
  ggplot(aes(x = BMI, y = LDL)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ x) +
  xlab("BMI (kg/m²)") +
  ylab("LDL cholesterol (mg/dL)") +
  theme_bw()
```

LDL cholesterol vs BMI in the HERS dataset, with linear regression fit

:::

#### Fit the model in R

```{r}
hers_lm_simple <- lm(
  LDL ~ BMI,
  data = hers,
  na.action = na.exclude
)
```

#### Parameter estimates

:::{#tbl-hers-simple-params}

```{r}
library(parameters)
hers_lm_simple |>
  parameters() |>
  print_md()
```

Parameter estimates for [Model @eq-hers-simple]

:::

#### Coefficient interpretation

::: notes
In [Model @eq-hers-simple]:
:::

-   $\beta_1$ is the difference in mean LDL cholesterol per
    1 kg/m² difference in BMI.
    From @tbl-hers-simple-params,
    $\hat\beta_1 \approx `r round(coef(hers_lm_simple)["BMI"], 2)`$ mg/dL
    per kg/m²,
    with a 95% confidence interval of
    `r paste(round(confint(hers_lm_simple)["BMI", ], 2), collapse = " to ")`
    mg/dL per kg/m².

-   $\beta_0$ is the mean LDL cholesterol for a woman with BMI = 0,
    which is outside the range of the data and therefore not directly
    interpretable.

### Multiple linear regression

::: notes
Next, we add potential confounders to the model,
following @vittinghoff2e Table 4.2.
:::

$$
\begin{aligned}
E[\text{LDL} \mid \vX = \vx] &= \beta_0 + \beta_{\text{BMI}} \cdot \text{BMI}
+ \beta_{\text{age}} \cdot \text{age}\\
&\quad + \beta_{\text{nw}} \cdot \text{nonwhite}
+ \beta_{\text{smk}} \cdot \text{smoking}\\
&\quad + \beta_{\text{drk}} \cdot \text{drinkany}
+ \beta_{\text{exr}} \cdot \text{exercise}
\end{aligned}
$$ {#eq-hers-multi}

#### Fit the model in R

```{r}
hers_lm_multi <- lm(
  LDL ~ BMI + age + nonwhite + smoking + drinkany + exercise,
  data = hers,
  na.action = na.exclude
)
```

#### Parameter estimates

:::{#tbl-hers-multi-params}

```{r}
hers_lm_multi |>
  parameters() |>
  print_md()
```

Parameter estimates for [Model @eq-hers-multi]

:::

#### Coefficient interpretation

::: notes
From @tbl-hers-multi-params:
:::

-   $\hat\beta_{\text{BMI}} \approx `r round(coef(hers_lm_multi)["BMI"], 2)`$
    mg/dL per kg/m²,
    after adjusting for age, race/ethnicity, smoking, alcohol use,
    and exercise.
    This is somewhat smaller than the unadjusted estimate
    ($\hat\beta_1 \approx `r round(coef(hers_lm_simple)["BMI"], 2)`$
    mg/dL per kg/m²) from [Model @eq-hers-simple],
    suggesting that some of the crude association between BMI and LDL
    is explained by the other covariates.

-   $\hat\beta_{\text{age}} \approx `r round(coef(hers_lm_multi)["age"], 2)`$
    mg/dL per year of age,
    adjusting for the other covariates.

### Comparing models

::: notes
To compare [Model @eq-hers-multi] and [Model @eq-hers-simple]
using a likelihood ratio test,
both models must be fit on the same observations.
We restrict to complete cases for all variables in the larger model.
:::

```{r}
library(dplyr)
hers_complete <- hers |>
  select(LDL, BMI, age, nonwhite, smoking, drinkany, exercise) |>
  na.omit()
```

```{r}
hers_lm_simple2 <- lm(
  LDL ~ BMI,
  data = hers_complete
)

hers_lm_multi2 <- lm(
  LDL ~ BMI + age + nonwhite + smoking + drinkany + exercise,
  data = hers_complete
)
```

```{r}
library(lmtest)
lrtest(hers_lm_multi2, hers_lm_simple2)
```

::: notes
The likelihood ratio test compares [Model @eq-hers-multi] to
[Model @eq-hers-simple].
A small p-value provides evidence that adding the additional
covariates improves model fit.
:::
