### Motivating example: `hers` data {.smaller}

:::{.callout-note}
This section is based on @vittinghoff2e, Chapter 4.
:::

::: notes

{{< include _sec_hers_intro.qmd >}}

:::

```{r}
#| eval: false
#| code-fold: show
library(haven)
hers <- haven::read_dta(
  paste0(
    "https://regression.ucsf.edu/sites/g/files",
    "/tkssra6706/f/wysiwyg/home/data/hersdata.dta"
  )
)
```

```{r}
#| include: false
library(haven)
hers <-
  fs::path_package("rme", "extdata/hersdata.dta") |>
  read_dta() |>
  dplyr::mutate(HT = as_factor(HT))
```

::::: {.panel-tabset}

#### Data as table

```{r}
#| label: tbl-hers-ch4
#| tbl-cap: "`hers` data"
hers |> head()
```

#### Data as graph

```{r}
#| label: fig-hers-scatter
#| fig-cap: "`hers` data (@vittinghoff2e)"
#| fig-height: 5
#| fig-width: 7
library(ggplot2)
hers_scatter <-
  hers |>
  ggplot(aes(
    x = BMI,
    y = LDL,
    shape = HT,
    col = HT
  )) +
  theme_bw() +
  xlab("BMI (kg/m²)") +
  ylab("LDL cholesterol (mg/dL)") +
  theme(legend.position = "bottom") +
  geom_point(alpha = .3)
print(hers_scatter + facet_wrap(~HT))
```

:::::

---

#### Data notation {.smaller}

::: notes
Let's define some notation to represent this data:
:::

-   $Y$: LDL cholesterol (mg/dL)
-   $T$: hormone therapy treatment assignment
    ("placebo" or "hormone therapy")
-   $H$: indicator variable for $T$ = "hormone therapy"
    -   $H = 0$ if $T$ = "placebo"
    -   $H = 1$ if $T$ = "hormone therapy"
-   $B$: BMI (kg/m²)

::: notes
"Placebo" is the **reference level** for the categorical variable $T$.
The choice of reference level is arbitrary;
it only affects the interpretation of the intercept and $\beta_H$.
:::

### Parallel lines regression for `hers` data {.smaller}

::: notes
We model the distribution of LDL cholesterol
as a linear function of BMI,
with a separate intercept for each treatment group:
:::

$$
\ba
Y|H,B &\sciid N(\mu(H,B), \sigma^2)\\
\mu(h,b) &= \beta_0 + \beta_H h + \beta_B b
\ea
$$ {#eq-hers-parallel}

::: {.column width=40%}

::: {#tbl-hers-lm1}

```{r}
library(dplyr)
hers_lm1 <- lm(
  formula = LDL ~ HT + BMI,
  data = hers,
  na.action = na.exclude
)

library(parameters)
hers_lm1 |>
  parameters::parameters() |>
  parameters::print_md(
    include_reference = include_reference_lines,
    select = "{estimate}"
  )
```

Regression parameter estimates for [Model @eq-hers-parallel]

:::

:::

:::{.column width=10%}
:::

:::{.column width=50%}

:::{#fig-hers-parallel-fit}

```{r}
hers <-
  hers |>
  dplyr::mutate(`E[LDL|X=x]` = fitted(hers_lm1)) |>
  dplyr::arrange(HT, BMI)

hers_scatter2 <-
  hers_scatter +
  geom_line(data = hers, aes(y = `E[LDL|X=x]`))

print(hers_scatter2)
```

Graph of [Model @eq-hers-parallel] for `hers` data

:::

:::
